---
title: "Using the interactive GTFS dashboard"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the interactive GTFS dashboard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`explore_gtfs()` provides an interactive Shiny dashboard for inspecting,
filtering, plotting, editing, and exporting a feed. `shiny` and `leaflet` are
optional dependencies; `plotly` is optional when interactive charts are
requested.

## Start the dashboard

Pass an existing object when working from R.

```{r, eval=FALSE}
library(GTFSwizard)
explore_gtfs(for_rail_gtfs)
```

In an interactive session, omitting the feed opens a file browser. This is the
simplest route for users who do not need to write R code.

```{r, eval=FALSE}
explore_gtfs()
```

Static `ggplot2` charts are the default. Enable Plotly only when hover and zoom
behavior is useful; the calendar and trip-duration boxplot remain static where
the conversion would reduce readability.

```{r, eval=FALSE}
explore_gtfs(for_rail_gtfs, plotly = TRUE)
```

## Filter the dashboard

The sidebar filters the working dashboard feed by routes, service patterns,
services, dates, time, and stops. Secondary choices are recomputed from the
current valid combination. An empty service selection means all services,
rather than no service.

Filters affect the maps, planning indicators, and service or performance plots.
Large route and service-pattern sets are summarized or limited in plots where
showing every class would make labels and legends unreadable. Plot-specific
`top_n` controls can expand those views when required.

## Interpret the planning indicators

The system and route tables summarize scheduled supply, not observed demand or
performance. Common indicators include route and stop counts, scheduled trips,
service span, vehicle-hours, peak fleet, total network length, and stop spacing.

Corridor share and hub share are proportions used to define which links or
stops are emphasized. Minimum corridor length is measured in meters. Adjust
these controls in the corridor and hub view; unsuitable or incomplete feeds
may not support every spatial analysis, in which case the dashboard leaves the
view unavailable and reports the reason.

## Edit without overwriting the source

The Edit tab applies `delay_trip()`, `split_trip()`, `edit_speed()`,
`set_dwelltime()`, or `edit_dwelltime()` to the dashboard's in-memory working
copy. The object or zip archive originally loaded is unchanged.

Review the affected trips and stops before applying an edit. Several edits can
be combined, and the Export GTFS control writes the currently filtered and
edited working feed to a destination selected by the user. A source file is
overwritten only when that exact path is deliberately chosen as the export
destination.

## Reproduce an analysis in R

The dashboard is useful for discovery. For a reproducible project, translate
the final choices into explicit function calls:

```{r, eval=FALSE}
scenario <- for_rail_gtfs |>
  filter_route(c("route-a", "route-b")) |>
  filter_time("06:00:00", "10:00:00") |>
  edit_speed(factor = 1.1)

write_gtfs(scenario, "morning-scenario.zip")
```

See [Filtering, selecting, and editing feeds](filtering-editing.html) for the
semantics of each operation.
